---
title: "EHR Project"
author: "Alisha Dziarski, Kari Guo & Susie Wang"
output: html_document
date: "`r Sys.Date()`"
---

```{r Setup}
library(tidyverse)
library(lubridate)
library(survival)
library(epiR)
library(epiDisplay)
library(cmprsk)
library(ggplot2)
library(lme4)
```

```{r Data Import}
hosp <- read.csv("hospitalisation_2018-2020.csv")
pres <- read.csv("prescription_2018-2020.csv")
diab_h <- read.csv("diabetes_hosp_2018-2020.csv")
diab_l <- read.csv("diabetes_lti_2018-2020.csv")
pat <- read.csv("tab_patient.csv")
```

```{r Data Cleaning}
## PATIENT tables -----------------------------------------------------

# Convert date columns to date format
pat$PAT_BIRTH_DATE <- date(pat$PAT_BIRTH_DATE)

# create placeholder death date for NAs to calculate age at end of study period
unique(pat$PAT_DEATH_DATE)   # 2020-12-31 is a death date so will set 2021-01-01 as placeholder

pat$PAT_DEATH_DATE <- date(ifelse(pat$PAT_DEATH_DATE == "",
                                  "2021-01-01", pat$PAT_DEATH_DATE))
dates <- pat$PAT_DEATH_DATE
dates[dates >= ymd("2021-01-01")] <- ymd("2021-01-01")
pat$PAT_DEATH_DATE <- dates

# Calculate age at death or age at end of study period
pat$PAT_AGE <- as.double(difftime(pat$PAT_DEATH_DATE, pat$PAT_BIRTH_DATE,
                                  units = "weeks")) / 52

# Status indicator at end of study (0 = alive, 1 = death)
pat$STATUS <- ifelse(pat$PAT_DEATH_DATE == "2021-01-01", 0, 1)

# Group by geographic area(Northeast, Northwest, Southeast, Southwest,
# Île-de-France, French territory islands (3-digits))
pat <- pat %>% add_column(PAT_DPT_GRP = NA, .after = "PAT_DPT_RES")

table(pat$PAT_DPT_RES)   # check unique values

pat$PAT_DPT_GRP <- ifelse(pat$PAT_DPT_RES %in% c("2", "8", "10", "21", "25",
                                                 "39", "51", "52", "54", "55",
                                                 "57", "59", "62", "67", "68",
                                                 "70", "71", "88", "90"), "NE",
       ifelse(pat$PAT_DPT_RES %in% c("14", "18", "22", "27", "28", "29", "35", "37",
                                     "41", "44", "45", "49", "50", "53", "56", "58",
                                     "60", "61", "72", "76", "80", "85"), "NW",
              ifelse(pat$PAT_DPT_RES %in% c("1", "3", "4", "5", "6", "7", "13", "15",
                                            "26", "30", "34", "38", "42", "43", "63",
                                            "66", "69", "73", "74", "83", "84"), "SE",
                     ifelse(pat$PAT_DPT_RES %in% c("9", "11", "12", "16", "17", "19",
                                                   "23", "24", "31", "32", "33",
                                                   "36", "40", "46", "47", "48",
                                                   "64", "65", "79", "81", "82",
                                                   "86", "87", "89", "209"), "SW",
                            ifelse(pat$PAT_DPT_RES %in% c("20", "75", "77", "78", "91",
                                                          "92", "93", "94", "95"), "IF",
                                   ifelse(pat$PAT_DPT_RES %in% c("0", "97", "971",
                                                                 "972", "973", "974",
                                                                 "978", "999"), "Other",
                                          "Unknown"))))))

# check which depts are still missing groupings
pat %>% filter(PAT_DPT_GRP == "Unknown") %>% 
  distinct(PAT_DPT_RES)


# Diabetic patients -- convert date columns to date format
diab_h$HOSP_START_DATE <- date(diab_h$HOSP_START_DATE)
diab_h$HOSP_END_DATE <- date(diab_h$HOSP_END_DATE)

diab_l$LTI_START_DATE <- date(diab_l$LTI_START_DATE)
diab_l$LTI_END_DATE <- date(diab_l$LTI_END_DATE)

# check
class(diab_h$HOSP_START_DATE)
class(diab_h$HOSP_END_DATE)
class(diab_l$LTI_START_DATE)
class(diab_l$LTI_END_DATE)


## PRESCRIPTION table -------------------------------------------------

# Convert date columns to date format
pres$PRS_DATE <- date(pres$PRS_DATE)

first_prs <- pres %>% group_by(PAT_ID) %>% 
  summarise("1ST_PRS_DATE" = min(PRS_DATE))

unique(first_prs$`1ST_PRS_DATE`)   # 1st date of Rx is same for all patients
pres$`1ST_PRS_DATE` <- date("2019-01-09")   # assign same date for 1st Rx to all patients


## HOSPITALIZATION table ----------------------------------------------

# Convert date columns to date format
hosp$HOSP_START_DATE <- date(hosp$HOSP_START_DATE)
hosp$HOSP_END_DATE <- date(hosp$HOSP_END_DATE)

# check
class(hosp$HOSP_START_DATE)
class(hosp$HOSP_END_DATE)

# Calculate hospitalization duration (in days)
hosp <- hosp %>% add_column(HOSP_DUR = NA, .after = "HOSP_END_DATE")
hosp$HOSP_DUR <- as.numeric(difftime(hosp$HOSP_END_DATE, hosp$HOSP_START_DATE,
                                     units = "days"))

# Calculate number of hospitalizations
hosp_freq <- hosp %>% group_by(PAT_ID, HOSP_START_DATE) %>%
  summarise(freq = frequency(HOSP_START_DATE))

hosp_sum <- hosp_freq %>% group_by(PAT_ID) %>% 
  summarise(NUM_HOSP = sum(freq))

hosp <- full_join(hosp, hosp_sum, by = "PAT_ID")


## COMBINING DATAFRAMES ----------------------------------------------

# Combine T2D patients from LTI and from hospitalizations
diab <- full_join(diab_h, diab_l, by = "PAT_ID")


# New dataframe for diabetes prescriptions
pres_diab <- pres %>% filter(str_detect(DRUG_ATC_C07, "A10A") |
                               str_detect(DRUG_ATC_C07, "A10B"))
table(pres_diab$DRUG_ATC_C07)   # double-check that A10As and A10Bs have been selected


# Join the diagnostic data with the prescription data by PAT_ID
full <- full_join(diab, pres_diab, by = "PAT_ID")
full$PAT_ID <- as.character(full$PAT_ID)

class(full)
dim(full)

# Clean out unnecessary columns 



```

## Logistic Regression
Logistic regression will then be performed with the explanatory variable being the four geographic areas, and the response variable being the binary outcome of the presence of at least one anti-diabetic drug prescription during the study period. An unadjusted model will be first conducted to investigate to understand an overall association between the geographic location of the patient and the presence of anti-diabetic drug prescription. A multivariate logistic regression will then be performed with age and sex as covariates to understand the impact of age and sex and confounders. The unadjusted and multivariate logistic models will then be compared to understand the impact of geographic location of diabetic patients on the prescription and intake of anti-diabetic drugs with and without the presence of confounders.  

```{r Hospitalization}
#Recode hospitalization into binary variable 
full$hospitalization <- full$HOSP_START_DATE
full$hospitalization <- ifelse(is.na(full$HOSP_START_DATE), 0, 1)

#Apply generalized linear model, currently using PAT_ID, but to replace with geographic area.
m1 <- glm(hospitalization~PAT_ID, data=full, family = "binomial")
summary(m1)

#Multivariate logistic regression
m2 <- glm(hospitalization~PAT_ID+age+sex, data=full, family="binomial")
summary(m2)
```


```{r Prescription}
#Recode prescription into binary variable 
full$prescription <- full$DRUG_NAME
full$prescription <- ifelse(is.na(full$prescription), 0, 1)

#Apply generalized linear model, currently using PAT_ID, but to replace with geographic area.
m3 <- glm(prescription~PAT_ID, data=full, family = "binomial")
summary(m3)

#Multivariate logistic regression
m4 <- glm(prescription~PAT_ID+age+sex, data=full, family="binomial")
summary(m4)
```

## Survival Analysis 
Aalen-Johansen estimation will be conducted to understand how the probabilities of anti-diabetic drug prescription and hospitalization change with time after DMII diagnosis. In both research questions, death will be the only competing risk considered. Cox proportional hazard models will then be performed for each research question. The hospitalization hazard among diabetic patients from different geographic areas of France will be compared with left truncation by age, considering hospitalizations within the study period. Additionally, a comparative time to first prescription of diabetes medication will be investigated with a Cox model left truncating by date of diagnosis. Assumptions of the Cox proportional hazard models will be checked to ensure the proper application of the models. The hazard ratios obtained, with their p-values and CIs will be interpreted and compared to better understand the impact of geographic residence of diabetic patients on the wait time for the initial prescription of anti-diabetic drugs and hospitalization, with age and sex potentially being the covariates.  

```{r Survival Analysis}
#Calculate time passed since DMII diagnosis for hospitalization 

#Calculate time passed since DMII diagnosis for prescription

#Fit Aalen-Johansen estimator for hospitalization
fit1 <- aareg(Surv(full$diagnosistime, full$status) ~ full$group) #group would be death or hospitalization

#Plot the cumulative incidence functions
plot(fit1, main = "Aalen-Johansen Cumulative Incidence Functions", col = c("red", "blue"))
legend("topright", legend = c("Death", "Hospitalization"), col = c("red", "blue"), lty = 1)

p1 <- ggsurvplot(fit = fit1, # Kaplan Meier survival model
                 data = full,
                 fun = "cumhaz", # "cumulative hazard" function
                 title = "'Survival Analysis' on the Prescription of Anti-diabetic Drugs and Hospitalization", #plot title
                 legend.labs = c("Prescription","Hospitalization"),    # Change legend labels
                 censor = F, # don't show censoring timepoints on the plot
                 xlab = "Time since DMII Diagnosis"
) 

p1
```

## Linear Regression 
A linear regression will be used to investigate the impact of geographic location of the diabetic patients on the secondary outcome, the number of hospitalizations. Potential confounders considered include age and sex.  

```{r Hospitalization}
#Calculate the number of hospitalizations 

#unadjusted linear model
l1 <- lm(num~area, full)
summary(l1)

#adjusted linear model 
l2 <- lm(num~area+age+sex, full)
summary(l2)
```

## Mixed Models 
Mixed model will be performed to understand the association between the number of hospitalization and the duration of each hospitalization stratified by each geographic area of the diabetic patients. Potential confounders include age, sex, and number of prescriptions.  

```{r Mixed Models}
#
m1 <- lmer(CD4~T+(1|NUM_PAT), albi, REML=0)
m1
summary(m1)

m2 <- lmer(CD4~T+(T|NUM_PAT), albi, REML=0)
m2
summary(m2)

# Run the mixed model for t that gives random intercept and random slope but no covariance between random intercept and random slope
m3 <- lmer(CD4~T+(1|NUM_PAT)+(0+T|NUM_PAT), albi, REML=0)
m3
summary(m3)

anova(m10)
```

## Sensitivity Analysis 

```{r Sensitivity Analysis}


```


