---
title: "EHR Project"
author: "Alisha Dziarski, Kari Guo & Susie Wang"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r Setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(gtsummary)
library(kableExtra)
library(epiR)
library(epiDisplay)
library(cmprsk)
library(ggplot2)
library(lme4)
```

```{r Data Import}
hosp <- read.csv("hospitalisation_2018-2020.csv")
pres <- read.csv("prescription_2018-2020.csv")
diab_h <- read.csv("diabetes_hosp_2018-2020.csv")
diab_l <- read.csv("diabetes_lti_2018-2020.csv")
pat <- read.csv("tab_patient.csv")
```

```{r Data Cleaning}
## PATIENT tables -----------------------------------------------------

# Convert date columns to date format
pat$PAT_BIRTH_DATE <- date(pat$PAT_BIRTH_DATE)

# create placeholder death date for NAs to calculate age at end of study period
unique(pat$PAT_DEATH_DATE)   # 2020-12-31 is a death date so will set 2021-01-01 as placeholder

pat$PAT_DEATH_DATE <- date(ifelse(pat$PAT_DEATH_DATE == "",
                                  "2021-01-01", pat$PAT_DEATH_DATE))

dates <- pat$PAT_DEATH_DATE
dates[dates >= ymd("2021-01-01")] <- ymd("2021-01-01")
pat$PAT_DEATH_DATE <- dates

# Status indicator at end of study (0 = alive, 1 = death)
pat$STATUS <- ifelse(pat$PAT_DEATH_DATE == "", 0, 1)

# Calculate age at death or age at end of study period
pat$PAT_AGE <- as.double(difftime(pat$PAT_DEATH_DATE, pat$PAT_BIRTH_DATE,
                                  units = "weeks")) / 52

# Group by geographic area(Northeast, Northwest, Southeast, Southwest,
# Île-de-France, French territory islands (3-digits))
pat <- pat %>% add_column(PAT_DPT_GRP = NA, .after = "PAT_DPT_RES")

table(pat$PAT_DPT_RES)   # check unique values

pat$PAT_DPT_GRP <- ifelse(pat$PAT_DPT_RES %in% c("2", "8", "10", "21", "25", "39",
                                                 "51", "52", "54", "55", "57",
                                                 "59", "62", "67", "68", "70",
                                                 "71", "88", "89", "90"), "NE",
       ifelse(pat$PAT_DPT_RES %in% c("14", "18", "22", "27", "28", "29", "35", "37",
                                     "41", "44", "45", "49", "50", "53", "56", "58",
                                     "60", "61", "72", "76", "80", "85"), "NW",
              ifelse(pat$PAT_DPT_RES %in% c("1", "3", "4", "5", "6", "7", "13", "15",
                                            "26", "30", "34", "38", "42", "43", "63",
                                            "66", "69", "73", "74", "83", "84"), "SE",
                     ifelse(pat$PAT_DPT_RES %in% c("9", "11", "12", "16", "17", "19",
                                                   "23", "24", "31", "32", "33",
                                                   "36", "40", "46", "47", "48",
                                                   "64", "65", "79", "81", "82",
                                                   "86", "87"), "SW",
                            ifelse(pat$PAT_DPT_RES %in% c("75", "77", "78", "91", "92",
                                                          "93", "94", "95"), "IF",
                                   ifelse(pat$PAT_DPT_RES %in% c("0", "20", "97",
                                                                 "98", "99", "209",
                                                                 "971", "972", "973",
                                                                 "974", "978", "999"),
                                          "Other", "Unknown"))))))

# check which depts are still missing groupings
pat %>% filter(PAT_DPT_GRP == "Unknown") %>% 
  distinct(PAT_DPT_RES)


# Diabetic patients -- convert date columns to date format
diab_h$HOSP_START_DATE <- date(diab_h$HOSP_START_DATE)
diab_h$HOSP_END_DATE <- date(diab_h$HOSP_END_DATE)

diab_l$LTI_START_DATE <- date(diab_l$LTI_START_DATE)
diab_l$LTI_END_DATE <- date(diab_l$LTI_END_DATE)

# check
class(diab_h$HOSP_START_DATE)
class(diab_h$HOSP_END_DATE)
class(diab_l$LTI_START_DATE)
class(diab_l$LTI_END_DATE)


## PRESCRIPTION table -------------------------------------------------

# Convert date columns to date format
pres$PRS_DATE <- date(pres$PRS_DATE)


## HOSPITALIZATION table ----------------------------------------------

# Convert date columns to date format
hosp$HOSP_START_DATE <- date(hosp$HOSP_START_DATE)
hosp$HOSP_END_DATE <- date(hosp$HOSP_END_DATE)

# check
class(hosp$HOSP_START_DATE)
class(hosp$HOSP_END_DATE)

# Calculate hospitalization duration (in days)
hosp <- hosp %>% add_column(HOSP_DUR = NA, .after = "HOSP_END_DATE")
hosp$HOSP_DUR <- as.numeric(difftime(hosp$HOSP_END_DATE, hosp$HOSP_START_DATE,
                                     units = "days"))

# Calculate number of hospitalizations
hosp_freq <- hosp %>% group_by(PAT_ID, HOSP_START_DATE) %>%
  summarise(freq = frequency(HOSP_START_DATE))

hosp_sum <- hosp_freq %>% group_by(PAT_ID) %>% 
  summarise(NUM_HOSP = sum(freq))

hosp <- full_join(hosp, hosp_sum, by = "PAT_ID")

## COMBINING DATAFRAMES ----------------------------------------------

# Combine T2D patients from LTI and from hospitalization with T2D
diab <- full_join(diab_h, diab_l, by = "PAT_ID")

## Clean up the dataframes -------------------------------------------
# Prescription table
pres <- pres %>% 
  dplyr::select(-c("DRUG_CIP7.1","DRUG_BIG_PACK","DRUG_FRM_GAL","DRUG_LAB_ID","PRS_HP_COD","PRS_NAT"))

# Hospitalization table
hops <- hosp %>% 
  dplyr::select(-c("ETA_NUM","RSA_NUM","HOSP_ENT_MODE","HOSP_ENT_ORIGIN","HOSP_EXIT_MODE","HOSP_EXIT_DEST"))

# Patient table 
pat <- pat %>% 
  dplyr::select(-c("PAT_HEA_INSUR","PAT_LOC_IRIS"))

# Diabetic Patient Table
diab <- diab %>% 
  dplyr::select(-c("ETA_NUM","RSA_NUM","HOSP_ENT_MODE","HOSP_ENT_ORIGIN","HOSP_EXIT_MODE","HOSP_EXIT_DEST","ETA_NUM.1","RSA_NUM.1"))

# Find the patient information that only has a diagbetic diagnosis 
diagnostic <- inner_join(pat, diab, by="PAT_ID")
dim(diagnostic)

# Join the prescription with patient data by PAT_ID
prescription <- full_join(diagnostic, pres, by = "PAT_ID")
prescription$PAT_ID <- as.character(prescription$PAT_ID)

class(prescription)
dim(prescription)

# Make sure all the data we need are in the correct format 
prescription$PAT_ID <- as.character(prescription$PAT_ID)
prescription$PAT_SEX_COD <- as.character(prescription$PAT_SEX_COD)
class(prescription$PAT_AGE)

class(diagnostic$PAT_SEX_COD)
diagnostic$PAT_ID<- as.character(diagnostic$PAT_ID)
diagnostic$PAT_SEX_COD <- as.character(diagnostic$PAT_SEX_COD)
```

```{r Table1}
## Using gtsummary
diagnostic %>% dplyr::select("PAT_DPT_GRP", "PAT_AGE", "PAT_SEX_COD") %>% 
  tbl_summary(by = "PAT_DPT_GRP")

diagnostic %>% group_by(PAT_DPT_GRP) %>% 
  summarize(minAge = min(PAT_AGE), maxAge = max(PAT_AGE))

## OR using kable
summary_table <- diagnostic %>%
  group_by(PAT_DPT_GRP) %>%
  summarize(Median_Age = median(PAT_AGE, na.rm = TRUE),
            Number_of_Males = sum(PAT_SEX_COD == 1, na.rm = TRUE),
            Percent_Males = (Number_of_Males / n()) * 100) %>%
  mutate('Number Males (%)' = paste0(Number_of_Males, " (",
                                     sprintf("%.1f%%", Percent_Males), ")")) %>%
  dplyr::select(-Number_of_Males, -Percent_Males)  # Drop intermediate columns

# Renaming columns
colnames(summary_table) <- c("Patient Location", "Median Age", "Percent Male (%)")

# Rename locations 
summary_table[1:6, 1] <- c("Ile de France", "Northeast", "Northwest",
                           "Other", "Southeast", "Southwest")

# Display the table using kable
print(kable(summary_table, caption = "Patient Characteristics Grouped by Location"))
```

## Logistic Regression
Logistic regression will then be performed with the explanatory variable being the four geographic areas, and the response variable being the binary outcome of the presence of at least one anti-diabetic drug prescription during the study period. An unadjusted model will be first conducted to investigate to understand an overall association between the geographic location of the patient and the presence of anti-diabetic drug prescription. A multivariate logistic regression will then be performed with age and sex as covariates to understand the impact of age and sex and confounders. The unadjusted and multivariate logistic models will then be compared to understand the impact of geographic location of diabetic patients on the prescription and intake of anti-diabetic drugs with and without the presence of confounders.  

```{r Hospitalization-Logistic}
#Recode hospitalization into binary variable 
diagnostic$hosp <- diagnostic$HOSP_START_DATE
diagnostic$hosp <- ifelse(is.na(diagnostic$HOSP_START_DATE), 0, 1)

#Apply generalized linear model
g1 <- glm(hosp~PAT_DPT_GRP, data=diagnostic, family = "binomial")
summary(g1)
r1 <- exp(g1$coefficients)
c1 <- exp(confint(g1))
cbind(OR=r1, c1)

#Multivariate logistic regression
g2 <- glm(hosp~PAT_DPT_GRP+PAT_AGE+PAT_SEX_COD, data=diagnostic, family="binomial")
summary(g2)
cbind(OR=exp(g2$coefficients),exp(confint(g2)))
```

```{r Prescription-Logistic}
#Recode prescription into binary variable 
prescription$pres <- prescription$DRUG_ATC_C07
prescription$pres <- ifelse(str_detect(prescription$DRUG_ATC_C07, "A10"), 1, 0)
prescription$pres <- ifelse(is.na(prescription$DRUG_NAME), 0, prescription$pres)

#Apply generalized linear model
g3 <- glm(pres~PAT_DPT_GRP, data=prescription, family = "binomial")
summary(g3)
cbind(OR=exp(g3$coefficients),exp(confint(g3)))

#Multivariate logistic regression
g4 <- glm(pres~PAT_DPT_GRP+PAT_AGE+PAT_SEX_COD, data=prescription, family="binomial")
summary(g4)
cbind(OR=exp(g4$coefficients),exp(confint(g4)))
```

## Linear Regression 
A linear regression will be used to investigate the impact of geographic location of the diabetic patients on the secondary outcome, the number of hospitalizations. Potential confounders considered include age and sex.  

```{r Hospitalization-Linear}
#Calculate the number of hospitalizations 
hosp$PAT_ID <- as.character(hosp$PAT_ID)
hospitalization <- inner_join(diagnostic, hosp, by="PAT_ID")

#unadjusted linear model
l1 <- lm(NUM_HOSP~PAT_DPT_GRP, data=hospitalization)
summary(l1)
cbind(Coefficient=l1$coefficient, confint(l1))

#adjusted linear model 
l2 <- lm(NUM_HOSP~PAT_DPT_GRP+PAT_AGE+PAT_SEX_COD, data=hospitalization)
summary(l2)
cbind(Coefficient=l2$coefficient, confint(l2))
```

## Mixed Models 
Mixed model will be performed to understand the association between the number of hospitalization and the duration of each hospitalization stratified by each geographic area of the diabetic patients. Potential confounders include age, sex, and number of prescriptions.  

```{r Mixed Models}
hospitalization$PAT_ID <- as.character(hospitalization$PAT_ID)

# Run the mixed model for different geographic departments only 
# only random intercept
m1 <- lmer(HOSP_DUR~NUM_HOSP+(1|PAT_ID), hospitalization, REML=0)
m1
summary(m1)

# random intercept + random slope
m2 <- lmer(HOSP_DUR~NUM_HOSP+(NUM_HOSP|PAT_ID), hospitalization, REML=0)
m2
summary(m2)

# Run the mixed model for the duration of hospitalization that gives random intercept and random slope but no covariance between random intercept and random slope
m3 <- lmer(HOSP_DUR~NUM_HOSP+(1|PAT_ID)+(0+NUM_HOSP|PAT_ID), hospitalization, REML=0)
m3
summary(m3)

# Comparing the 3 models to decide the best model 
anova(m1,m2)
anova(m1,m3)
anova(m2,m3)

# From the result, we don't see lots of differences among the 3 models. Therefore, after considering the necessity of random slope, we decided to simplify the model to only include the random intercept. 
confint(m1)
```

## Sensitivity Analysis 

```{r sensitivity analysis}
# Adding the geographic departments to the equation to see if there is any change in the results. 
m4 <- lmer(HOSP_DUR~NUM_HOSP+PAT_DPT_GRP+(1|PAT_ID), hospitalization, REML=0)
m4
summary(m4)
confint(m4)

# With all potential confounders 
m5 <- lmer(HOSP_DUR~PAT_DPT_GRP+NUM_HOSP+PAT_AGE+PAT_SEX_COD+(1|PAT_ID), hospitalization, REML=0)
m5
summary(m5)
confint(m5)
```